\thetitle

### 2017-07-18 introduction.tex
starting to restructure the text

\subsection*{Essentially algebraic theories}
An essentially algebraic theory is an equational theory of partial operations
which may be given the structure of a presheaf over usual ordered set of the
natural numbers, and of which the domain of each operation is declared by 
equations on the preceeding operations. The concept of essentially algebraic
theories is due to Freyd, who writes that a critical feature of essentially
algebraic theories is that their models are closed under 
colimits \cite{Aspects_of_topoi}. Many classes of theories have been described
and shown to be equivalent to essentially algebraic theories. Among them are
Partial Horn Logic and quasi-equational theories \cite{Palmgren_Vickers},
cartesian theories \cite{Freyd2002}, finite limit sketches 
\cite{Toposes_triples_theories}, left exact theories
\cite{McLarty1986}, and they can be equivalently represented as locally
presentable categories \cite{Adamek_Locally}.

@@

\subsection*{Setting up the essentially algebraic theory of E-systems}
The first goal of this paper is to describe an essentially algebraic theory of
type dependency that is in close correspondence to the type theory of
\cite{TheBook}, while at the same time it is as economic as possible in the
demand of requirements of the meta-theory (such as the availability of a type of natural numbers). Instead of explaning
what a dependent type is via an infinite hierarchy of universes $\UU_0$,
$\UU_1$, \ldots, we describe type dependency using only the three sorts
that typically occur in type theoretical expressions: contexts, dependent contexts and 
terms. Although an infinite hierarchy of universes allows one to reduce everything
to a term (because every type is a term of some specified universe), and 
therefore use the literally same mechanisms for handling type expressions as for
term expressions in the process of compilation and verification of a 
construction, this specific setup requires the natural numbers to exist in the 
meta-theory.

In the specification of a formal language of type theory, we would therefore
have three pairs of kinds of judgments that can be made:
\begin{align*}
\jalign\jctx{\Gamma} 
& \jalign\jctxeq{\Gamma}{\Gamma'}
  \\
\jalign\jfam{\Gamma}{A} 
& \jalign\jfameq{\Gamma}{A}{B}
  \\
\jalign\jterm{\Gamma}{A}{x} 
& \jalign\jtermeq{\Gamma}{A}{x}{y}.
\end{align*}
These are asserting that $\Gamma$ is a context; that $\Gamma$ and $\Gamma'$ are
judgementally equal contexts; that $A$ is a family in context $\Gamma$; that
$A$ and $B$ are judgmentally equal families in context $\Gamma$; that $x$ is
a term of $A$ in context $\Gamma$; and that $x$ and $y$ are judgmentally equal
terms of $A$ in context $\Gamma$.

a E-system is therefore an algebra consisting of three sorts: $C$ for contexts,
$F$ for families, and $T$ for terms. There is a map 
$\eft:F\to C$ which determines for each family, in which context it is a
family; and there is a map $\ebd:T\to F$ which determines for each term, of
which family it is a term. Thus, the backbones of a E-system can be represented
as
\begin{equation*}
\begin{tikzcd}
T \arrow[d,"\ebd"] \\ F \arrow[d,"\eft"] \\ C
\end{tikzcd}
\end{equation*}
This structure is represented by the following set of rules governing judgmental
equality. First, judgmental equality is postulated to be an equivalence relation
in all three cases (contexts, families and terms), so we require
%\bgroup\small
\begin{align*}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\Gamma}{\Gamma}
    } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      }
    { \jctxeq{\Delta}{\Gamma}
      } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jctxeq{\Delta}{\greek{E}}
      }
    { \jctxeq{\Gamma}{\greek{E}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{A}{A}
    } 
& & \inference
    { \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Gamma}{B}{A}
      }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jfameq{\Gamma}{B}{C}
      }
    { \jfameq{\Gamma}{A}{C}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jtermeq{\Gamma}{A}{x}{x}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{A}{y}{x}
      }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      \jtermeq{\Gamma}{A}{y}{z}
      }
    { \jtermeq{\Gamma}{A}{x}{z}
      }
\end{align*}
%\egroup
Second, there are convertibility rules assuring that the notion of being a
family in a certain context, or a term of a family, is compatible with judgmental
equality.
\begin{align*}
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jfam{\Gamma}{A}
    }
  { \jfam{\Delta}{A}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Delta}{A}{B}
      }
    \\
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Delta}{A}{x}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Delta}{A}{x}{y}
      }
    \\
& \inference
  { \jfameq{\Gamma}{A}{B}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Gamma}{B}{x}
    }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{B}{x}{y}
      }
\end{align*}
When $A$ is a family in context $\Gamma$, there is also a context $\ctxext{\Gamma}{A}$,
called the \emph{context extension of $A$}. Context extension allows us to consider
`families over families', where a family $P$ over $A$ in context $\Gamma$, would
simply be a family $P$ over $\ctxext{\Gamma}{A}$. Context extension is represented
by another function $\cftctxext : F\to C$.

@@

The category of B-systems will be shown to be equivalent to the category of
stratified E-systems. However, we do not arrive at the fact that the functor
$\mathcal{E}:\mathbf{Bsys}\to\mathbf{Ctx}$ is full without imposing an additional
condition on the E-homomorphisms. 
To ensure that the category of stratified E-systems with stratified 
E-homomorphisms between them is a full subcategory of the category 
$\mathbf{Ctx}$, we will require that the underlying functor of each E-homomorphism
lifts factorizations. It is
important to note that this condition is also algebraic, so that the category
of E-systems remains algebraic.

Recall that for any morphism $f:X\to Y$ in a category $\cat{C}$ there is a category
$\mathbf{fact}(f)$ of factorizations of $f$. Also, any functor $F:\cat{C}\to\cat{D}$
determines a functor $F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(H(f))$
for any morphism $f$ of $\cat{C}$.

\begin{defn}
A functor $F:\cat{C}\to\cat{D}$ is said to
\define{lift factorizations} if for any $f\in\cat{C}/X$, the functor
$F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(F(f))$ is an isomorphism
of categories. 
\end{defn}

\begin{comment}
\begin{rmk}
We may choose the property of lifting factorizations
to involve either an equivalence or an isomorphism of categories. The
isomorphism-version, would say that for every factorization $h'\circ g'$
of $F(f)$ there is a unique factorization $f\jdeq h\circ g$ in $\cat{C}$ such
that $F(g)\jdeq g'$ and $F(h)\jdeq h'$. In the version with equivalences,
the uniqueness is replaced by uniqueness up to isomorphism, and the equalites
are replaced by an isomorphism in $\mathbf{fact}(F(f))$.

@@

\subsection*{Overview and conventions}
In \autoref{sec:esys_defn}, we give the categorical definition of E-systems in 
several stages. First we define weakening systems, then projection systems and
substitution systems, and finally we bundle them together into the definition
of E-systems. The structure of the definition of weakening systems is 
prototypcial: we first define the notion of pre-weakening systems, which are
categories that come with a pre-weakening structure; then we define the notion
of pre-weakening homomorphism, which is a functor preserving the weakening
structure and we note that it makes sense to ask for the property that the
functors of which the weakening system consists, are also pre-weakening
homomorphisms; and finally, adding this requirement will be give the definition
of weakening systems. We go through this process for each of the definitions
of weakening systems, projection systems, substitution systems and E-systems.

@@

\begin{rmk}
When we think of a category $\cat{F}$ as a category of fibrations or families
over a context -- which will be the outset of the definition of E-systems --
it will be convenient to denote the objects of the $\cat{F}$ with capital
greek letters $\Gamma$, $\Delta$, \ldots, and the morphisms by capital latin letters
$A$, $B$, $P$, $Q$, \ldots. When $A$ is a morphism into $\Gamma$, we will denote
its domain by $\ctxext{\Gamma}{A}$, so we don't always have to introduce a name
for the domain. When $A$ is a morphism into $\Gamma$ and
$P$ is a morphism into $\ctxext{\Gamma}{A}$, it will also be convenient to denote the
composition of $P$ with $A$ by $\ctxext{A}{P}$. Note that $\ctxext{A}{P}$
is indeed the domain of $P$ in the slice category $\cat{F}/\Gamma$.
\end{rmk}

@@

\subsection*{Acknowledgements}
Several people have been indispensible throughout the research process.
I would like to thank 
Richard Garner for the early discussions on the inference rules of type 
dependency we had in Barcelona and in Paris; 
Joachim Kock for the many fruitful discussions and for hosting me at the Universitat
Aut\`onoma de Barcelona during the year 2013-2014;
Steve Awodey for sharing many insights and useful advice which lead to the
definition of a E-system in arbitrary categories with finite limits;
and Vladimir Voevodsky for inviting me to present E-systems at the Homotopy Type
Theory workshop in Oxford (November 2014) and to visit the Institute for Advanced 
Study in Princeton (March 2015), where the definition presented here was
discovered.

>>>>

\subsection*{Essentially algebraic theories}
An essentially algebraic theory is an equational theory of partial operations
which may be given the structure of a presheaf over usual ordered set of the
natural numbers, and of which the domain of each operation is declared by 
equations on the preceding operations. The concept of essentially algebraic
theories is due to Freyd, who writes that a critical feature of essentially
algebraic theories is that their models are closed under 
colimits \cite{Aspects_of_topoi}. Many classes of theories have been described
and shown to be equivalent to essentially algebraic theories. Among them are
Partial Horn Logic and quasi-equational theories \cite{Palmgren_Vickers},
cartesian theories \cite{Freyd2002}, finite limit sketches 
\cite{Toposes_triples_theories}, left exact theories
\cite{McLarty1986}, and they can be equivalently represented as locally
presentable categories \cite{Adamek_Locally}.

@@

\subsection*{Setting up the essentially algebraic theory of E-systems}
The first goal of this paper is to describe an essentially algebraic theory of
type dependency that is in close correspondence to the type theory of
\cite{TheBook}, while at the same time it is as economic as possible in the
demand of requirements of the meta-theory (such as the availability of a type of natural numbers). Instead of explaining
what a dependent type is via an infinite hierarchy of universes $\UU_0$,
$\UU_1$, \ldots, we describe type dependency using only the three sorts
that typically occur in type theoretical expressions: contexts, dependent contexts and 
terms. Although an infinite hierarchy of universes allows one to reduce everything
to a term (because every type is a term of some specified universe), and 
therefore use the literally same mechanisms for handling type expressions as for
term expressions in the process of compilation and verification of a 
construction, this specific setup requires the natural numbers to exist in the 
meta-theory.

In the specification of a formal language of type theory, we would therefore
have three pairs of kinds of judgments that can be made:
\begin{align*}
\jalign\jctx{\Gamma} 
& \jalign\jctxeq{\Gamma}{\Gamma'}
  \\
\jalign\jfam{\Gamma}{A} 
& \jalign\jfameq{\Gamma}{A}{B}
  \\
\jalign\jterm{\Gamma}{A}{x} 
& \jalign\jtermeq{\Gamma}{A}{x}{y}.
\end{align*}
These are asserting that $\Gamma$ is a context; that $\Gamma$ and $\Gamma'$ are
judgmentally equal contexts; that $A$ is a family in context $\Gamma$; that
$A$ and $B$ are judgmentally equal families in context $\Gamma$; that $x$ is
a term of $A$ in context $\Gamma$; and that $x$ and $y$ are judgmentally equal
terms of $A$ in context $\Gamma$.

An E-system is therefore going to be defined as an algebra consisting of three sorts: $C$ for contexts,
$F$ for families, and $T$ for terms. There is a map 
$\eft:F\to C$ which determines for each family, in which context it is a
family; and there is a map $\ebd:T\to F$ which determines for each term, of
which family it is a term. Thus, an E-system is a structure on top of the following simpler structure:
\begin{equation*}
\begin{tikzcd}
T \arrow[d,"\ebd"] \\ F \arrow[d,"\eft"] \\ C
\end{tikzcd}
\end{equation*}
This structure is represented by the following set of rules governing judgmental
equality. First, judgmental equality is postulated to be an equivalence relation
in all three cases (contexts, families and terms), so we require
%\bgroup\small
\begin{align*}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\Gamma}{\Gamma}
    } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      }
    { \jctxeq{\Delta}{\Gamma}
      } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jctxeq{\Delta}{\greek{E}}
      }
    { \jctxeq{\Gamma}{\greek{E}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{A}{A}
    } 
& & \inference
    { \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Gamma}{B}{A}
      }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jfameq{\Gamma}{B}{C}
      }
    { \jfameq{\Gamma}{A}{C}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jtermeq{\Gamma}{A}{x}{x}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{A}{y}{x}
      }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      \jtermeq{\Gamma}{A}{y}{z}
      }
    { \jtermeq{\Gamma}{A}{x}{z}
      }
\end{align*}
%\egroup
Second, there are convertibility rules assuring that the notion of being a
family in a certain context, or a term of a family, is compatible with judgmental
equality.
\begin{align*}
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jfam{\Gamma}{A}
    }
  { \jfam{\Delta}{A}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Delta}{A}{B}
      }
    \\
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Delta}{A}{x}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Delta}{A}{x}{y}
      }
    \\
& \inference
  { \jfameq{\Gamma}{A}{B}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Gamma}{B}{x}
    }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{B}{x}{y}
      }
\end{align*}
When $A$ is a family in context $\Gamma$, there is also a context $\ctxext{\Gamma}{A}$,
called the \emph{context extension of $A$}. Context extension allows us to consider
`families over families', where a family $P$ over $A$ in context $\Gamma$, would
simply be a family $P$ over $\ctxext{\Gamma}{A}$. Context extension is represented
by another function $\cftctxext : F\to C$.

@@

The category of B-systems will be shown to be equivalent to the category of
stratified E-systems. However, we do not arrive at the fact that the functor
$\mathcal{E}:\mathbf{Bsys}\to\mathbf{Esys}$ is full without imposing an additional
condition on the E-homomorphisms. 
To ensure that the category of stratified E-systems with stratified 
E-homomorphisms between them is a full subcategory of the category 
$\mathbf{Esys}$, we will require that the underlying functor of each E-homomorphism
lifts factorizations. It is
important to note that this condition is also algebraic, so that the category
of E-systems remains algebraic.

Recall that for any morphism $f:X\to Y$ in a pre-category $\cat{C}$ there is a pre-category
$\mathbf{fact}(f)$ of factorizations of $f$. Also, any functor $F:\cat{C}\to\cat{D}$
determines a functor $F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(H(f))$
for any morphism $f$ of $\cat{C}$.

\begin{defn}\label{defn:lift_factorizations}
A functor $F:\cat{C}\to\cat{D}$ is said to
\define{lift factorizations} if for any $f\in\cat{C}/X$, the functor
$F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(F(f))$ is an isomorphism
of pre-categories. 
\end{defn}

\begin{comment}
\begin{rmk}
We may choose the property of lifting factorizations
to involve either an equivalence or an isomorphism of categories. The
isomorphism-version, would say that for every factorization $h'\circ g'$
of $F(f)$ there is a unique factorization $f\jdeq h\circ g$ in $\cat{C}$ such
that $F(g)\jdeq g'$ and $F(h)\jdeq h'$. In the version with equivalences,
the uniqueness is replaced by uniqueness up to isomorphism, and the equalities
are replaced by an isomorphism in $\mathbf{fact}(F(f))$.

@@

\subsection*{Overview and conventions}
In \autoref{sec:esys_defn}, we give the categorical definition of E-systems in 
several stages. First we define weakening systems, then projection systems and
substitution systems, and finally we bundle them together into the definition
of E-systems. The structure of the definition of weakening systems is 
prototypical: we first define the notion of pre-weakening systems, which are
categories that come with a pre-weakening structure; then we define the notion
of pre-weakening homomorphism, which is a functor preserving the weakening
structure and we note that it makes sense to ask for the property that the
functors of which the weakening system consists, are also pre-weakening
homomorphisms; and finally, adding this requirement will be give the definition
of weakening systems. We go through this process for each of the definitions
of weakening systems, projection systems, substitution systems and E-systems.

@@

\begin{rmk}
When we think of a category $\cat{F}$ as a category of fibrations or families
over a context -- which will be the outset of the definition of E-systems --
it will be convenient to denote the objects of the $\cat{F}$ with capital
Greek letters $\Gamma$, $\Delta$, \ldots, and the morphisms by capital Latin letters
$A$, $B$, $P$, $Q$, \ldots. When $A$ is a morphism into $\Gamma$, we will denote
its domain by $\ctxext{\Gamma}{A}$, so we don't always have to introduce a name
for the domain. When $A$ is a morphism into $\Gamma$ and
$P$ is a morphism into $\ctxext{\Gamma}{A}$, it will also be convenient to denote the
composition of $P$ with $A$ by $\ctxext{A}{P}$. Note that $\ctxext{A}{P}$
is indeed the domain of $P$ in the slice category $\cat{F}/\Gamma$.
\end{rmk}

@@

\subsection*{Acknowledgements}
Several people have been indispensable throughout the research process.
I would like to thank 
Richard Garner for the early discussions on the inference rules of type 
dependency we had in Barcelona and in Paris; 
Joachim Kock for the many fruitful discussions and for hosting me at the Universitat
Aut\`onoma de Barcelona during the year 2013-2014;
Steve Awodey for sharing many insights and useful advice which lead to the
definition of a E-system in arbitrary categories with finite limits;
and Vladimir Voevodsky for inviting me to present E-systems at the Homotopy Type
Theory workshop in Oxford (November 2014) and to visit the Institute for Advanced 
Study in Princeton (March 2015), where the definition presented here was
discovered.
